---
title: "Propensity Score Update"
author: "Jamaal Green"
date: "1/2/2018"
output:
  pdf_document: default
  word_document: default
  md_document:  default
bibliography: dissertation_Oct2017.bib
---

```{r setup, include=FALSE, cache = TRUE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE)



if(!require(pacman)){install.packages("pacman"); library(pacman)}
p_load(RPostgreSQL, stargazer, texreg, pander, lmtest, multiwayvcov, cobalt, ipw,  dplyr, dbplyr)


#logistic table processing from industrial_land db for MANUFACTURING JOBS ONLY------
host <- "pgsql.rc.pdx.edu"
user <- "jamgreen"
pw <- "&zSpR-rmd&v5REgZ"
dbname <- "industrial_land"


con <- dbConnect("PostgreSQL", host = host, user = user, dbname = dbname, password = pw)




city_lehd_mfg <- dbGetQuery(con, "select distinct a.blk_grp_id, a.tot_emp, a.ag_emp, a.mining_emp,
a.util_emp, a.mfg_emp, a.wholesale_emp, a.transpo_emp,
a.place_geoid, a.city_name, b.totpop as totpop, b.white_nh as white_nh, 
b.black_nh as black_nh, b.hispanic as hispanic, b.mhi as mhi, 
b.housing_units as tot_units, b.owner_occ as owner_occ, 
b.renter_occ as renter_occ,
d.d3a as network_density, 
b.totpop/(ST_Area(a.geom)/1000000) as pop_density,
c.pmd_dummy as pmd_dummy, a.geom
	FROM lehd_places_2004_nad83albers a 
		LEFT JOIN propensity_blkgrp b ON
		a.blk_grp_id = b.geoid 
		LEFT JOIN lehd04_pmd c ON
		a.blk_grp_id = c.blk_grp_id
		LEFT JOIN epa_sld d ON
		a.blk_grp_id = d.geoid10;")



city_lehd_mfg <- city_lehd_mfg %>% 
  mutate(ind_emp = ag_emp + mining_emp + util_emp + mfg_emp + wholesale_emp + transpo_emp,
  IndShare = ind_emp/tot_emp, MfgShare = mfg_emp/tot_emp) %>% 
  distinct(blk_grp_id, .keep_all = TRUE)


#filter out blockgroups with greater than 34 tot_emp (first quartile) and replace NAs with 0
#city_lehd_mfg <- city_lehd_dist %>% filter(tot_emp > 34)
city_lehd_mfg[is.na(city_lehd_mfg)] <- 0

#calculate population shares for white and non-white
city_lehd_mfg <- city_lehd_mfg %>% mutate(BlackPer = black_nh/totpop, HispPer = hispanic/totpop,
                                          RenterPer = renter_occ/tot_units)

city_lehd_mfg$BlackPer[is.nan(city_lehd_mfg$BlackPer)] <- 0
city_lehd_mfg$HispPer[is.nan(city_lehd_mfg$HispPer)] <- 0
city_lehd_mfg$RenterPer[is.nan(city_lehd_mfg$RenterPer)] <- 0

#join to the city_dist table for final calcs

city_dist <- tbl(con, "city_dist_km")
city_dist <- collect(city_dist)

city_lehd_mfg <- city_lehd_mfg %>% left_join(city_dist, by = c("blk_grp_id" = "blk_grp_id")) %>% 
  distinct(blk_grp_id, .keep_all = TRUE)

city_lehd_mfg$pmd <- ifelse(city_lehd_mfg$pmd_dummy == TRUE, 1, 0)

#using ipw to compare the weights calculation-----
temp_mfg_stabilized <- ipwpoint(exposure = pmd_dummy, family = "binomial", link = "logit",
                                numerator = ~ 1, 
                                denominator = ~ MfgShare + BlackPer + HispPer + RenterPer + dist_km + network_density,
                                data = city_lehd_mfg)

temp_mfg_unstabilized <- ipwpoint(exposure = pmd_dummy, family = "binomial", link = "logit",
                                  denominator = ~ MfgShare + BlackPer + HispPer + RenterPer + dist_km +    network_density,
                                  data = city_lehd_mfg)


temp_ind_stable <- ipwpoint(exposure = pmd_dummy, family = "binomial", link = "logit",
                            numerator = ~ 1, 
                            denominator = ~ IndShare + BlackPer + HispPer + RenterPer + dist_km,
                            data = city_lehd_mfg)

temp_ind_unstable <- ipwpoint(exposure = pmd_dummy, family = "binomial", link = "logit",
                              denominator = ~ IndShare + BlackPer + HispPer + RenterPer + dist_km + network_density,
                              data = city_lehd_mfg)

city_lehd_mfg$ipw_ind_stable <- temp_ind_stable$ipw.weights
city_lehd_mfg$ipw_ind_unstable <- temp_ind_unstable$ipw.weights

city_lehd_mfg$ipw_mfg_stable <- temp_mfg_stabilized$ipw.weights
city_lehd_mfg$ipw_mfg_unstable <- temp_mfg_unstabilized$ipw.weights

#weighted regression models, have to query the db to get 2009 and 2015 emp numbers------
#**first steps are to create the final model table----

prop_2009 <- tbl(con, "lehd_model_2009_emp")
prop_2009 <- collect(prop_2009)
prop_2015 <- tbl(con, "lehd_model_2015_emp")
prop_2015 <- collect(prop_2015)

prop_final <- prop_2015 %>% select(1, 12:17) %>% inner_join(prop_2009, by = "geoid10")

prop_final <- prop_final %>% mutate(mfg_change = mfg2015 - mfg2009,
            ind_2009 = ag2009 + mining2009 + util2009 + mfg2009 + wholesale2009 + transpo2009,
            ind_2015 = ag2015 + mining2015 + util2015 + mfg2015 + wholesale2015 + transpo2015,
            ind_change = ind_2015 - ind_2009)

prop_final <- prop_final %>% select(geoid10, city = city_name, bg_fips,
                                    pmd_dummy, mfg_ps, ipw_mfg_stable, ipw_mfg_unstable,
                                    ind_ps, ipw_ind_stable, ipw_ind_unstable,
                                    mfg_change, ind_change)
```

## Introduction

In order to test the effectiveness of PMD-style land-use policies I collected block group level employment data from the LEHD for the years 2004, 2009, and 2015 for 47 of the country's 50 largest cities (three cities were dropped due to data availability issues of the LEHD) and 2005-2009 ACS data for the initial logistic model in order to calculate propensity weights. What follows is a description of the process thus far and some outstanding issues regarding the propensity score model specification. 

##Data

The unit of analysis is the census block group- a census geography that holds between 600 and 3000 people. Within the 47 cities there are 33,985 block groups with either employment only data from the LEHD, population only data from the ACS, or some combination therein. From the LEHD I collected total employment, manufacturing and "industrial" employment (the sum of agricultural, mining, utilities, transportation/warehousing, wholesale trade, and manufacturing). From the ACS I collected total population, Non-Hispanic Whites, Non-Hispanic Blacks, Hispanics, the total number of housing units, and owner/renter occupied units (all race/ethnicity data from single race reports). From the EPA Smart Location Database I collected the street network density of each block group. Finally, I calculated the Euclidean distance from the city center to each block group's centroid in kilometers.

While a hefty sample, the 33,995 block groups are still primarily residential districts with minimal employment. As such, the distribution for jobs within cities are extremely left skewed with most block groups having zero employment or only a few jobs. This is to be expected given the primarily residential nature of most neighborhoods and that manufacturing/industrial employment is usually concentrated in specific block groups due to zoning restrictions.  

| Industry  | Min. | 1st Qu. | Median |  Mean | 3rd Qu. |   Max  |
|-----------|:----:|:-------:|:------:|:-----:|:-------:|:------:|
| Mfg. Emp  |   0  |    0    |    0   | 56.91 |    6    | 25,528 |
| Ind. Emp  |   0  |    0    |    4   | 133.1 |    28   | 57,130 |
| Total Emp |   0  |    34   |   119  | 690.4 |   403   | 202,442 |

Table 1: Summary Employment Statistics for Full Dataset

Of the 33,985 block groups only 155 of them are covered by a industrial preservation policy or 0.4% of all block groups.The 0.45% of block groups covered by an industrial preservation policy have a  higher number of all jobs, manufacturing, and industrial jobs than non-covered block groups.



##Propensity Score Model


In order to accurately estimate the effect of the preservation policies on manufacturing and industrial employment I am using a propensity score approach [@AustinIntroductionPropensityScore2011; @Rosenbaumcentralrolepropensity1983]. Propensity scores are the probability of a unit of observation receiving treatment conditioned on a set of characteristics. Traditionally, researchers use a logistic or probit regression in order to estimate the probabilities and then calculate weights using the probabilities. For this project I am using the inverse probability method with the goal of estimating the Average Treatment Effect (ATE) which is the average effect of moving the entire population from control to treatment (in this case transitioning block groups to industrial preservation). Once the propensity score is estimated and the weights calculated the weights are used in a weighted regression of the treatment indicator variable regressed on the dependent variable. If the indicator variable is significant, then this signals a causal effect of the treatment on your observation. 

In order to calculate weights I estimated an initial logistic regression calculating predicted probability of a block group being covered by an industrial preservation policy:

$$E(y|x) = \beta_0 + \beta_1 x_1... + \beta_k x_k$$

The regressors here are share of employment that is manufacturing or industrial, percent of the block group population that is Non-Hispanic Black, percent of the block group that is Hispanic, percent of block group's housing units that are renter occupied, distance from the city central business district to the block group centroid in kilometers, and street network density. 

| Variable                   |    Year    | Data Source |
|----------------------------|:----------:|------------:|
| % Black                    | 2005-2009  |         ACS |
| % Hispanic                 | 2005-2009  |         ACS |
| % Manufacturing/Industrial |    2004    |        LEHD |
| Distance from CBD (km)     |      -     |  Calculated |
| Street Network Density     |    2010    |     EPA SLD |

Table 2: Independent variables for the logit model 

###Unstabilized and Stabilized Propensity Weights

There are two main approaches for calculating propensity weights once the predicted probabilities are calculated- stabilized and unstabilized weights. The equation for unstabilized inverse probability weights for observations in treatment group is $\omega = 1/\hat{e}(x)$ and for the control group $\omega = 1/1-\hat{e}(x)$. While often used, unstabilized weights can have very wide distributions of weights and can hamper overall model efficiency. Stabilized weights add a term to the numerator, in the case of this model the numerator is the probability of the exposure level, in this case it is simply an intercept only logistic regression on the treatment variable- $s\omega_i = \beta_0/E(y|x)$. The following models use both stabilized and unstabilized weights calculated by the ipw package in the R Statistical Programming Lanugage version 3.4.2.

The following logitics regression tables show the underlying models that will be used to calculate the eventual weights. All dependent variables are significant and follow their expected signs. Table 4 shows the moderating influence of standardizing weights significantly shrinking the size and distribution of weights for the manufacturing and industrial employment models. 

 

```{r echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

require(stargazer)

stargazer(temp_mfg_unstabilized$den.mod, temp_ind_stable$den.mod, title = "Logit Models for The Probability of Treatment for Manufacturing and Industrial Jobs", type = "latex", column.labels = c("Manufacturing", "Industry"), header = FALSE)

```

\pagebreak

| Weight Label      |  Min. | 1st Qu. | Median | Mean  | 3rd Qu. | Max      |
|-------------------|:-----:|--------:|--------|-------|---------|----------|
| Mfg. Unstabilized |  1.00 |   1.002 | 1.003  | 1.849 | 1.005   | 1587.922 |
| Mfg. Stabilized   | 0.023 |   0.997 | 0.998  | 0.999 | 1.000   | 7.242    |
| Ind. Unstabilized | 1.000 |   1.001 | 1.002  | 1.696 | 1.004   | 1377.630 |
| Ind. Stabilized   | 0.022 |   0.996 | 0.997  | 0.999 | 0.999   | 6.283    |

Table 4: Stabilized and Unstabilized Propensity Score Weights for Industrial and Manufacturing Employment Models


While the stabilized and usntabilized weights give a sense of the different distributions of weights across models we are more interested in whether the weights properly balance model covariates. In order to assess the accuracy of the weighting I calculated the standardized mean differences for the model's main covariates using R's cobalt package [@GreifercobaltCovariateBalance2018]. Generally, standardized differences should be as close to 0 as possible and conservative measures place an upper limit between .1 and .25. Using these measures all variables in the propensity score  model except for the share of employment in manufacturing or industrial employment are appropriately balanced. While not perfect, these balance estimates do imply a reasonably balanced dataset and bias reduction.


```{r echo=FALSE, message=FALSE, warning=FALSE}

require(dplyr)
require(pander)
require(cobalt)

unstable_mfg <- city_lehd_mfg %>% select(MfgShare, RenterPer, BlackPer,
                                         HispPer, network_density, dist_km)

mfg_bal <- bal.tab(unstable_mfg, treat = city_lehd_mfg$pmd, weights = city_lehd_mfg$ipw_mfg_unstable)

myvars <- c("RenterPer", "BlackPer", "HispPer", "network_density", "dist_km", "IndShare")
unstable_ind <- city_lehd_mfg[myvars]

ind_bal <- bal.tab(unstable_ind, treat = city_lehd_mfg$pmd, weights = city_lehd_mfg$ipw_ind_stable)

ind_tbl <- tibble::tribble(~Variable,	~`Std. Mean Diff.`,
"IndShare",	0.7545,
"RenterPer",	-0.0813,
"BlackPer",	0.0158,
"HispPer",	0.0666,
"network_density",	-0.4115,
"dist_km",	0.209) 

mfg_tbl <- tibble::tribble(~Variable,	~`Std. Mean Diff.`,
"MfgShare",	0.4259,
"RenterPer",	-0.0035,
"BlackPer",	-0.0119,
"HispPer",	0.0974,
"network_density",	0.0312,
"dist_km",	0.0004
)

pander(mfg_tbl, justify = c("left", "right"), caption = "Standardized Mean Difference Scores for the Manufacturing Model")

pander(ind_tbl, justify = c("left", "right"), caption = "Standardized Mean Difference Scores for the Industrial Model")

```


##Final Model Results

###Manufacturing

In order to evaluate the effect of an industrial preservation policy, I estimated a propensity score regression regressing the change in manufacturing employment from 2009-2015 on whether or not a block group is covered by an industrial land use preservation policy. Ultimately, industrial preservation policies do not have statistically significant effects on the change in manufacturing or industrial employment between 2009 and 2015 for any model using the stabilized weights or that used clustered standard errors. 

Table 7 presents the final propensity score models for manufacturing employment change using standard OLS estimation and clustered robust standard errors on both stabilized and unstabilized weights. Using standard OLS estimations show a positive and significant averate treatment effect (ATE) of approximately 14.2 jobs. This means that if all block groups were to be treated we would expect to see a gain of 14.2 jobs between 2009 and 2015. Correcting for the clustered nature of the data shows a non-significant, though positive, effect of an industrial preservation policy.[^1] 


```{r echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

require(texreg)
#unstable manufacturing

unstable_mfg1 <- lm(mfg_change ~ pmd_dummy, weights = mfg_ps, data = prop_final)

mfg_unstable_cluster <- multiwayvcov::cluster.vcov(unstable_mfg1, cluster = prop_final$city)
cf_mfg1 <- lmtest::coeftest(unstable_mfg1, mfg_unstable_cluster)

stable_mfg1 <-lm(mfg_change ~ pmd_dummy, weights = ipw_mfg_stable, data = prop_final)

mfg_stable_cluster <- multiwayvcov::cluster.vcov(stable_mfg1, cluster = prop_final$city)
cf_stable_mfg1 <- lmtest::coeftest(stable_mfg1, mfg_stable_cluster)


#unstable_mfg2 <- lm(mfg_change ~ pmd_dummy + factor(city), weights = mfg_ps,data = prop_final)


# Adjust standard errors https://www.jakeruss.com/cheatsheets/stargazer/
#unstable_mfg1_vcov    <- vcovCR(unstable_mfg1, type = "CR1", cluster = "city")
#unstable_mfg1_robust_se    <- sqrt(diag(unstable_mfg1_vcov))

# adjust f-test
#unstable_mfg1_wald_results <- waldtest(unstable_mfg1, vcov = unstable_mfg1_vcov)

#stargazer(unstable_mfg1, lmtest::coeftest(unstable_mfg1, mfg_unstable_cluster), type = "latex", header = FALSE, #title = "Change in Manufacturing Employment with Unstabilized Weights")                             

texreg(list(unstable_mfg1, cf_mfg1, stable_mfg1, cf_stable_mfg1), 
       custom.model.names = c("OLS Unstable", "CRSE Unstable", "OLS Stabilized", "CRSE Stabilized"),
       label = "tab:5", 
       caption = "Propensity Score Model of Change in Manufacturing Employment (2009-2015) with Stabilized and Unstabilized Weights", caption.above = TRUE)
```


###Industrial

Industrial employment, a composite variable made up of the sum of agriculatural, mining, utilities, warehousing, transportation, and manufacturing employment benefitted from industrial land use preservation. Industrial land preservation policies result in a positive and significant change in industrial employment across all model specifications. Accroding to the final propensity scored model the ATE of the industrial preservation policy is a little over 176.5 between 2009 and 2015.


```{r echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

#unstable and stabilized industrial

unstable_ind1 <- lm(ind_change ~ pmd_dummy, weights = ind_ps, data = prop_final)

ind_unstable_cluster <- multiwayvcov::cluster.vcov(unstable_ind1, cluster = prop_final$city)
cf_ind1 <- lmtest::coeftest(unstable_ind1, ind_unstable_cluster)

stable_ind1 <-lm(ind_change ~ pmd_dummy, weights = ipw_ind_stable, data = prop_final)

ind_stable_cluster <- multiwayvcov::cluster.vcov(stable_ind1, cluster = prop_final$city)
cf_stable_ind1 <- lmtest::coeftest(stable_ind1, ind_stable_cluster)


texreg(list(unstable_ind1, cf_ind1, stable_ind1, cf_stable_ind1), 
       custom.model.names = c("OLS Unstable", "CRSE Unstable", "OLS Stabilized", "CRSE Stabilized"),
       label = "tab:5", 
       caption = "Propensity Score Model of Change in Industrial Employment (2009-2015) with Stabilized and Unstabilized Weights", caption.above = TRUE)
```

##Discussion

Given the model results industrial land preservation policies had a non-significant effect on the change in manufacturing employment form 2009-2015 but such policies did have a positive and significant effect for the broader category of industrial jobs. This positive and significant effect was maintained after correcting for potential bias by clustering standard errors at the city level. The model specification demonstrates a causal link between industrial land preservation policies and the protection of industrial jobs through the Great Recession and the first couple of years of recovery. Endogeneity concerns were addressed by using inverse probability weights calculated using a propensity score approach. 

[^1]: Tables were produced using the stargazer package [@MarkHlavacStargazer2015]


#References



<!---Running the first logistic model I received an error that my full model (manufacturing share, % Black, % Hispanic, % Renter, distance from city center, population density, total manufacturing employment) exhibited complete separation. This means that one of my covariates perfectly predicts my dependent variable. This generally results in unnaturally large regression coefficients and larger standard errors.--->