---
title: "LEHD Urban Industry Broad Patterns"
author: "Jamaal Green"
date: "10/31/2017"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Update on Methodology

I've collected LEHD data from 2004 to 2015 for the 48 cities where LEHD data is available. What follows is a brief cover of the industrial employment patterns for these cities over the study period. For now, these graphics are aspatial, though I will attempt a few select summary maps if space allows.

```{r message=FALSE, warning=FALSE, include=FALSE}
#bring in LEHD table and appropriate packages

getwd()

if(!require(pacman)){install.packages("pacman"); library(pacman)}
p_load(ggplot2, ggthemes, extrafont, hrbrthemes,viridis, sf, dplyr)

extrafont::font_import()
extrafont::loadfonts(quiet = TRUE)

lehd <- readRDS("~/Diss_Mfg_Modeling/data/spatial/lehd_citiesNAD83.rds")
city_growth <- lehd %>% as.data.frame() %>%  group_by(NAME, year) %>% 
  summarise(TotEmp = sum(C000), MfgEmp = sum(CNS05), 
            IndEmp = sum(CNS02 + CNS03 + CNS05 + CNS06 + CNS08))

city_growth <- city_growth %>% group_by(NAME) %>%  
  mutate(MfgBase = first(MfgEmp), IndBase = first(IndEmp),
         MfgIndex = (MfgEmp - MfgBase)/MfgBase * 100,
         IndIndex = (IndEmp - IndBase)/IndBase * 100) %>% ungroup()

city_growth$MfgPer <- city_growth$MfgEmp/city_growth$TotEmp

outlier_cities <- city_growth %>% filter(year == "2015") %>% 
  mutate(MfgIndexZ = scale(MfgIndex),
         IndIndexZ = scale(IndIndex)) %>% 
  select(NAME, MfgIndexZ, IndIndexZ)

city_growth <- left_join(city_growth, outlier_cities)

city_growth <- city_growth %>% 
  mutate(OutlierDummy1 = ifelse(MfgIndexZ > 1, "High", ifelse(MfgIndexZ < -1, "Low", "Average")),
         OutlierDummy2 = ifelse(IndIndexZ > 1, "High", ifelse(IndIndexZ < -1, "Low", "Average")))

city_growth$OutlierDummy1 <- as.factor(city_growth$OutlierDummy1)
levels(city_growth$OutlierDummy1) <- c("Low", "Average", "High")

city_growth$OutlierDummy2 <- as.factor(city_growth$OutlierDummy2)
levels(city_growth$OutlierDummy2) <- c("Low", "Average", "High")

city_growth <- city_growth %>% mutate(Policy = ifelse(NAME %in% c("Los Angeles", "Chicago", "San Diego", "Jacksonville",
                                                                  "San Francisco", "Boston", "Seattle", "Baltimore", "Portland",
                                                                  "San Jose", "Minneapolis", "New York"), 1, 0))
```

##Manufacturing Is Still Here Though It Is Weak

An initial look at this data shows a few things but an important one is that manufacturing still has a legitimate presence in city labor markets in both absolute and relative terms and that the state of manufacturing across cities is decidedly mixed. This first graphic, while admittedly busy, shows the broad pattern of the change in manufacturing employment over time in the nation's largest cities. Most cities, unfortunately, have lost manufacturing employment compared to 2005, corresponding to a more general belief that manufacturing is losing in the country and that cities, especially, are inhospitable places. But there are exceptions. The cities highlighted in red all had positive manufacturing employment growth compared to the blue cities that suffered catstrophic manufacturing job losses.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

mfg_index_plot <- ggplot(city_growth, aes(year, MfgIndex, group = NAME, color = OutlierDummy1)) + 
  geom_line() +  
  scale_color_manual(values = c("#DCDCDC", "#FF0000", "#0000FF")) +
  theme_ipsum_rc(grid = "Y", plot_title_size = 10) +
  labs(title = "Change in Mfg Employment for 48 Large Cities, 2005-2015",
       subtitle = "Mfg. Employment Indexed to 2005 measures",
       x = "Year", y = "Mfg. Employment Index") +
  theme(legend.position = "none")

mfg_index_plot
```

Six cities- Houston, Arlington, San Jose, Oklahoma City, Raleigh, and Tulsa- are the red lines in the previous graphic. By far the most dramatic growth in the past decade are in the energy centers of Oklahoma City, Houston, and Arlington. All three have greater than 10% employment growth in manufacturing when compared to their 2005 job numbers and Houston's manufacturing employment grew by nearly 25%. Though these cities have impressive manufacturing job growth, the *relative* importance of manufacturing employment weakened in many of the cities as the proportion of all jobs in manufacturing fell. 

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

growth_cities <- city_growth %>% filter(year == "2005" | year == "2015", NAME == "Houston" | NAME == "Arlington" | NAME == "Oklahoma City"|
                                          NAME == "Raleigh" | NAME == "Tulsa" | NAME == "San Jose") %>% 
  select(NAME, Year = year, TotEmp, MfgEmp)

growth_p1 <- ggplot(data = growth_cities, aes(x = NAME, y = MfgEmp, fill = Year)) + 
  geom_bar(stat = "identity", position = position_dodge()) + theme_ipsum_rc(plot_title_size = 10) +
  labs(title = "Cities Leading the Country in Manufacturing Growth",
       x = "City Name", y = "Manufacturing Employment") + scale_y_comma()

growth_p1

growth_cities <- growth_cities %>% ungroup() %>% mutate(MfgShare = MfgEmp/TotEmp)

growth_p2 <- ggplot(data = growth_cities, aes(x = NAME, y = MfgShare, fill = Year)) + 
  geom_bar(stat = "identity", position = position_dodge()) + theme_ipsum_rc(plot_title_size = 10) +
  labs(title = "Mfg. Employment is Less Important in Cities Where It's Growing",
       subtitle = "Cities with robust growth saw Mfg.'s share of total employment fall",
       x = "City Name", y = "Manufacturing Employment") + scale_y_percent()

growth_p2

```

Note that San Jose is the only city with an industrial land use preservation policy, but it also, by far, is the most dependent on manufacturing. We can see in this graphic that most cities with some sort of protective policy have relatively little manufacturing employment and have been *losing* manufacturing employment as a share of all employment over the decade. As the second indexed graphic shows, every protective city, except for San Jose, has consistently lost manufacturing employment since 2005, and in some cases, in dramatic fashion. 

If there is one great shared trait of cities experimenting with industrial land use preservation it would seem to be that they are already on the losing side of deindustrialization, though admittedly, a few of the cities are seeing an uptick in recent manufacturing job growth though not yet at 2005 levels. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

policy_cities <- city_growth %>% filter(Policy == 1)

policy_p1 <- ggplot(data = policy_cities, aes(year, MfgPer)) + geom_bar(stat = "identity", fill = "steelblue") +
  theme_ipsum_rc(plot_title_size = 14, strip_text_face = "bold", strip_text_size = 8) + 
  facet_wrap(~NAME, scales = "free", ncol = 4) + scale_y_percent() +
  labs(title = "Mfg. Employment Share for Protective Cities", 
       x = "Year", y = "Mfg. Employment Share (%)",
       caption = "Boston is not included due to lack of available LEHD data")

policy_p1
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
policy_index_plot <- ggplot(policy_cities, aes(year, MfgIndex, group = NAME)) + 
  geom_line() +  
  theme_ipsum_rc(grid = "Y", plot_title_size = 10) +
  labs(title = "Change in Mfg Employment for Protective Cities, 2005-2015",
       subtitle = "Mfg. Employment Indexed to 2005 measures",
       x = "Year", y = "Mfg. Employment Index") +
  theme(legend.position = "none")

policy_index_plot

```